\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-05-08 fluB.tex
Change citation style and text/scripts related to LD.
Changed citation style to MBE-like. Added more information about D' and
why it's horrible for flu sequences.
Also improved efficiency of LD calculator. LD results will be derived
from ~1600 complete influenza B genomes and will stick with statistics
related to r-squared. Might implement minor allele frequency cutoff
though.

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

@@

\section*{Introduction}
Seasonal influenza causes between 250,000 and 500,000 deaths annually and comprises lineages from three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \cite{flufactsheet}.
However, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.

Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments, a process known as reassortment. 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to their ability to cause pandemics in humans via reassortment of circulating human strains with non-human influenza A strains. 
Although influenza B viruses have been observed to infect seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis, they are thought to primarily infect humans and are thus are unlikely to exhibit pandemics due to the absence of an animal reservoir from which to acquire antigenic novelty. 
Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{burnet1955,hay2001,bedford2014}.

Currently circulating influenza B viruses comprise two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/2/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \cite{rota1990}. 
These two lineages now possess antigenically distinct HA surface glycoproteins \cite{kanegae1990,rota1990,nerome1998,nakagawa2002,ansaldi2003} allowing them to co-circulate in the human population.
Phylogenetic analysis of evolutionary rate, selective pressures and reassortment history of influenza B has shown extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within the Vic and Yam lineages \cite{chen2008}.

@@

We show the ratio of Vic and Yam lineages in our sequence data in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on lineage assignment performed earlier.
It is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to the repeated fixation of either the Vic or the Yam lineage.
These losses of diversity correspond to fixation of the Vic lineage in NS and fixation of the Yam lineage in PA, NP, NA and MP.
Similarly, the lack of reassortment between Vic and Yam lineages and maintenance of diversity of PB1, PB2 and HA can be seen, where the two lineages have been isolated at a ratio close to 50\% over long periods of time (Figure \ref{lineageRatiosOverTime}).
On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically consistent with one lineage predominating within a given season, in agreement with surveillance data \cite{reed2012}.

@@

We reconstructed reassortment events that were detected by using lineage labels.
Figure \ref{railroadPlot} focuses only on inter-lineage reassortments that have occurred after 1990.
We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS with prime (') indicating independently acquired segments) circulating between 1992 and 2011 (Figure \ref{railroadPlot}):
\begin{itemize}
  \item B/Alaska/12/1996-like (Y-Y-Y-Y-Y-Y-Y-V)
  \item B/Nanchang/2/1997-like (V-V-Y-V-Y-V-Y-V)
  \item B/Iowa/03/2002-like (V-V-Y'-V-Y-Y-Y'-V')
  \item B/California/NHRC0001/2006-like (V-V-Y-V-Y'-Y-Y'-V')
  \item B/Brisbane/33/2008-like (V-V-Y-V-Y'-Y-Y-V)
\end{itemize}
B/Alaska/12/1996-like, B/Nanchang/2/1997-like and B/Iowa/03/2002-like constellations were observed by Chen \& Holmes \cite{chen2008}, but sequences from B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like constellations were not available at the time.
Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata derived background.
All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata derived PB1-PB2-HA segments.
Figure \ref{railroadPlot} also shows that reassorting segments appear to evolve with a considerable degree of autonomy.
For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to the rise of viruses with B/Iowa/03/2002-like genome constellations.
A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has originally been derived from the Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
Overall, we see that successful inter-lineage reassortment events do not break up Vic and Yam PB1-PB2-HA complexes, but any other segments reassorted into one or the other PB1-PB2-HA background evolve with a considerable degree of autonomy.

@@

\subsection*{Analysis of within-lineage reassortment patterns}
SPR (subtree prune and regraft) provides a measure of tree-to-tree distance, in which distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between them \cite{svinti2013}, as long as each branch experiences, at most, one reassortment.
Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
If there are biases in the way segments reassort, so that some segments tend to co-assort more often, we expect to observe a lower reassortment rate between them, which would manifest as small-scale similarities between phylogenetic trees of those segments.
In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the overall (\textit{i.e} both within and between lineages) reassortment rate.

@@

We normalized our SPR distance comparisons to account for differences in tree topology stability over the course of the MCMC chain and also use an approximation of SPR distance \cite{whidden2009,whidden2010,whidden2013} (see Methods).
The 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments between segments (Figure \ref{SPRdistances}).
Reassortment rate given as number of SPR moves per total time in both trees shows similar results (Figure \ref{NormSPR_RErate}).
We note, however, that because of phylogenetic uncertainty our estimate of SPR distance might simply lack power.
Comparisons between independent analyses of the same segments yield distances that are comparable to distances between different segments (Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
Still, we find that independent replicates from the same segment (Figure \ref{SPRdistancesReplicates}) show lower SPR distances that comparisons between segments (Figure \ref{SPRdistancesTrees}), suggesting that phylogenetic noise is not completely overwhelming reassortment signal.
In addition, SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
Thus we caution against over-interpreting Figure \ref{SPRdistances}.
Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_difference}).

Although we do not find evidence of differences in numbers of overall reassortments between segments, we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
Unlike SPR distances, which are counts of the numbers of reassortment events, $\dtmrca$ takes time into account, and is most sensitive when only one of the two trees being compared loses diversity via reassortment.
Similarly to SPR distances, we normalized our $\dtmrca$ comparisons to account for node date stability over the course of the MCMC chain (see Methods).
Figure \ref{deltaTMRCA} shows normalized mean $\dtmrca$ values for all pairs of trees.
Most segment pairs show very low values for this statistic with $\dtmrca\approx$0.1, indicating that $\dtmrca$ measurements between independent analyses of the same segment are up to 10 times smaller than $\dtmrca$ values between analyses of different segments.
PB1, PB2 and HA trees, on the other hand, exhibit normalized $\dtmrca$ values that are much higher.
This shows that TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occasionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
For example, the distribution of comparison between PB1 and PB2 trees has a mean $\dtmrca$$\approx0.4$, indicating that the sum of $\dtmrca$ values from PB1-PB1' and PB2-PB2' comparisons (prime (') indicates an independent MCMC chain) are 40\% of twice the $\dtmrca$ value from the PB1-PB2 comparison, but this value can range anywhere between $\approx$0.15 and $\approx$0.75.
Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort among themselves unless both reassorting segments have similar TMRCAs, i.e.\ we observe isolation by distance.
In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.

@@

Linkage disequilibrium (LD) is a measure of non-random association between pairs of alleles at different polymorphic loci within a population.
To estimate LD no other information, except for allele frequencies at polymorphic amino acid sites, is required.
Although not an external validation of previous phylogenetic results, we observe significantly greater LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}).
This suggests that PB1, PB2 and HA segments possess a considerable number of co-assorting alleles, which upon closer inspection are associated with either Vic or Yam lineage segments.
We conclude that Victoria and Yamagata lineages of PB1, PB2 and HA have accumulated lineage-specific amino acid substitutions.
Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
Figure \ref{ChiGenome} also shows that the number of variable amino acid sites in PB1 and PB2 proteins is comparable to most other proteins investigated (see Methods for cut-offs used to exclude sites from LD analysis).
These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
In addition, there are sites on PB1, PB2 and HA proteins which also show high, albeit smaller, LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.

@@

When comparing the mean tip-tip TMRCA deviations we find evidence of isolation by distance - when reassortments occur between PB1, PB2 and HA segments those events tend to involve branches that have similar TMRCAs.
Other segment pairs, with the exception of MP-NA, seem to show little signal of this and are able to reassort with branches that have been evolving under temporally distant genomic backgrounds.
In addition, most strains with mixed-lineage PB1-PB2-HA complexes occurred in the early years of the Vic--Yam split, when the two lineages were presumably much more similar at the nucleotide and amino acid levels.
The most recent influenza B viruses with B/Waikato/6/2005-like genome constellations possessed mixed-lineage PB1-PB2-HA complexes and circulated for only 5 months before presumably going extinct.

@@

\subsection*{Causes of co-dependence between PB1-PB2-HA}
Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, \textit{e.g.} PB1-PB2-PA and HA-NA \cite{mccullers2004}.
Though it would be easy to explain co-adaptation between these segments by referring to their functional roles \textit{e.g.} PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities, our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.

It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together: there have been at least 4 sampled mixed-lineage PB1-PB2-HA complex constellations which did not become fixed in the population and our estimates of the number of reassortments do not differ significantly between all segments (Figure \ref{SPRdistances}).
The latter is in line with recent experiments in influenza A that have shown that reassortment between segments differing by a single synonymous difference is highly efficient \cite{marshall2013}.
In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we offer several alternatives.

@@

Explaining the co-dependence of PB1+2 and HA segments is more difficult.
Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \cite{bergeron2010,fulvini2011}.
A previously used technique for producing vaccine seed strains involved selecting for HA-NA reassortants, which often resulted in PB1-HA-NA reassortants, the second most frequent class of reassortants \cite{bergeron2010,fulvini2011}.
Recent experiments have suggested that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \cite{cobbin2013}.
Additional evidence for a relationship between PB1 and HA segments in influenza A viruses is given by previous influenza pandemics, which were caused by avian-human influenza A virus reassortants.
It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments derived from avian influenza A viruses \cite{kawaoka1989}.
However, there have also been reassortant influenza A viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \cite{gregory2002} and H1N1/09 in 2009 \cite{smith2009}.

@@

Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
This is easiest to interpret as frequency-dependent balancing selection acting to preserve antigenic diversity \cite{worobey2014}.
If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.

It is possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
In this case PB1, PB2 and HA segments of Vic and Yam lineage accumulate substitutions that improve their ability to co-operate with segments of the same lineage and worse at interacting segments of a different lineage.
This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \cite{lee2008}.
It is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \cite{nerome1998,shaw2002}, offering a potential explanation for why the budding Victoria lineage segments were not homogenized via reassortment with Yamagata lineage segment in the early years of the split between the two lineages. 
The restriction to eastern Asia would presumably also give sufficient time for PB1, PB2 and HA segments to co-evolve together.
Similarly to the previous scenario, however, mutation-driven co-evolution would require an association of some kind between PB1+2 and HA segments to explain the low observed frequency and poor fitness of PB1+2/HA reassortants.
In addition, we see no reason why these three particular segments, and not the rest of the genome, would become co-adapted to each other in geographic isolation.

\subsection*{Suggested experiments}
The association between Victoria and Yamagata lineage PB1-PB2-HA complexes should be relatively straightforward to test in the lab.
Using previously developed plasmid systems \cite{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
We predict that artificially produced viruses with mixed-lineage PB1-PB2-HA complexes will have reduced fitness when compared to viruses with pure-lineage, i.e.\ entirely Vic or entirely Yam, PB1-PB2-HA complexes.
In addition, we expect the relationship between Vic and Yam lineage PB1, PB2 and HA segments to be dependent on date of segment isolation, as viruses with mixed-lineage PB1-PB2-HA complexes isolated earlier should perform better than viruses with PB1-PB2-HA segments isolated more recently.

@@

However, we also see that epistatic effects might interfere with fitness measurements, if for example non-PB1-PB2-HA segments are also temporally mismatched.
Ideally, the co-adaptation would be easier to understand by referring to the structures of PB1 and PB2 proteins, as the link between these would be intuitive.
We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins (Figure \ref{ChiCore}) that fall within the regions that form contacts within the influenza B polymerase heterotrimer \cite{sugiyama2009}, suggesting more subtle roles for sites we have identified.

>>>>

% cite package, to clean up citations in the main text. Do not remove.
%\usepackage{cite}
\usepackage{natbib}
%\renewcommand\citepleft{(}
%\renewcommand\citepright{)}
%\renewcommand\citepform[1]{\textsl{#1}}

\usepackage{amsmath}

@@

\section*{Introduction}
Seasonal influenza causes between 250,000 and 500,000 deaths annually and comprises lineages from three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \citep{flufactsheet}.
However, influenza B viruses are increasingly being recognized as important human pathogens \citep{paul-glezen2013}.
Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \citep{ECDC1213}.

Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments, a process known as reassortment. 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to their ability to cause pandemics in humans via reassortment of circulating human strains with non-human influenza A strains. 
Although influenza B viruses have been observed to infect seals \citep{osterhaus2000,bodewes2013} through a reverse zoonosis, they are thought to primarily infect humans and are thus are unlikely to exhibit pandemics due to the absence of an animal reservoir from which to acquire antigenic novelty. 
Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \citep{burnet1955,hay2001,bedford2014}.

Currently circulating influenza B viruses comprise two distinct lineages -- Victoria and Yamagata (referred to as Vic and Yam, respectively) -- named after strains B/Victoria/2/87 and B/Yamagata/16/88, that are thought to have genetically diverged in HA around 1983 \citep{rota1990}. 
These two lineages now possess antigenically distinct HA surface glycoproteins \citep{kanegae1990,rota1990,nerome1998,nakagawa2002,ansaldi2003} allowing them to co-circulate in the human population.
Phylogenetic analysis of evolutionary rate, selective pressures and reassortment history of influenza B has shown extensive and often complicated patterns of reassortment between all segments of influenza B viruses both between and within the Vic and Yam lineages \citep{chen2008}.

@@

We show the ratio of Vic and Yam lineages in our sequence data in different influenza seasons in Figure \ref{lineageRatiosOverTime}, which is based on lineage assignment performed earlier.
It is evident that losses of diversity in the PA, NP, NA, MP and NS segments are related to the repeated fixation of either the Vic or the Yam lineage.
These losses of diversity correspond to fixation of the Vic lineage in NS and fixation of the Yam lineage in PA, NP, NA and MP.
Similarly, the lack of reassortment between Vic and Yam lineages and maintenance of diversity of PB1, PB2 and HA can be seen, where the two lineages have been isolated at a ratio close to 50\% over long periods of time (Figure \ref{lineageRatiosOverTime}).
On a year-to-year basis, however, the ratios for Vic and Yam lineage PB1, PB2 and HA can fluctuate dramatically consistent with one lineage predominating within a given season, in agreement with surveillance data \citep{reed2012}.

@@

We reconstructed reassortment events that were detected by using lineage labels.
Figure \ref{railroadPlot} focuses only on inter-lineage reassortments that have occurred after 1990.
We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS with prime (') indicating independently acquired segments) circulating between 1992 and 2011 (Figure \ref{railroadPlot}):
\begin{itemize}
  \item B/Alaska/12/1996-like (Y-Y-Y-Y-Y-Y-Y-V)
  \item B/Nanchang/2/1997-like (V-V-Y-V-Y-V-Y-V)
  \item B/Iowa/03/2002-like (V-V-Y'-V-Y-Y-Y'-V')
  \item B/California/NHRC0001/2006-like (V-V-Y-V-Y'-Y-Y'-V')
  \item B/Brisbane/33/2008-like (V-V-Y-V-Y'-Y-Y-V)
\end{itemize}
B/Alaska/12/1996-like, B/Nanchang/2/1997-like and B/Iowa/03/2002-like constellations were observed by Chen \& Holmes \citep{chen2008}, but sequences from \\B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like constellations were not available at the time.
Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata derived background.
All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata derived PB1-PB2-HA segments.
Figure \ref{railroadPlot} also shows that reassorting segments appear to evolve with a considerable degree of autonomy.
For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to the rise of viruses with B/Iowa/03/2002-like genome constellations.
A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has originally been derived from the Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.
Overall, we see that successful inter-lineage reassortment events do not break up Vic and Yam PB1-PB2-HA complexes, but any other segments reassorted into one or the other PB1-PB2-HA background evolve with a considerable degree of autonomy.

@@

\subsection*{Analysis of within-lineage reassortment patterns}
SPR (subtree prune and regraft) provides a measure of tree-to-tree distance, in which distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between them \citep{svinti2013}, as long as each branch experiences, at most, one reassortment.
Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
If there are biases in the way segments reassort, so that some segments tend to co-assort more often, we expect to observe a lower reassortment rate between them, which would manifest as small-scale similarities between phylogenetic trees of those segments.
In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the overall (\textit{i.e} both within and between lineages) reassortment rate.

@@

We normalized our SPR distance comparisons to account for differences in tree topology stability over the course of the MCMC chain and also use an approximation of SPR distance \citep{whidden2009,whidden2010,whidden2013} (see Methods).
The 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments between segments (Figure \ref{SPRdistances}).
Reassortment rate given as number of SPR moves per total time in both trees shows similar results (Figure \ref{NormSPR_RErate}).
We note, however, that because of phylogenetic uncertainty our estimate of SPR distance might simply lack power.
Comparisons between independent analyses of the same segments yield distances that are comparable to distances between different segments (Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
Still, we find that independent replicates from the same segment (Figure \ref{SPRdistancesReplicates}) show lower SPR distances that comparisons between segments (Figure \ref{SPRdistancesTrees}), suggesting that phylogenetic noise is not completely overwhelming reassortment signal.
In addition, SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
Thus we caution against over-interpreting Figure \ref{SPRdistances}.
Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_difference}).

Although we do not find evidence of differences in numbers of overall reassortments between segments, we find support for a reassortment `distance' effect, in which a pair of tips on one segment has a different TMRCA from the same pair of tips on a different segment.
Unlike SPR distances, which are counts of the numbers of reassortment events, $\dtmrca$ takes time into account, and is most sensitive when only one of the two trees being compared loses diversity via reassortment.
Similarly to SPR distances, we normalized our $\dtmrca$ comparisons to account for node date stability over the course of the MCMC chain (see Methods).
Figure \ref{deltaTMRCA} shows normalized mean $\dtmrca$ values for all pairs of trees.
Most segment pairs show very low values for this statistic with $\dtmrca\approx$0.1, indicating that $\dtmrca$ measurements between independent analyses of the same segment are up to 10 times smaller than $\dtmrca$ values between analyses of different segments.
PB1, PB2 and HA trees, on the other hand, exhibit normalized $\dtmrca$ values that are much higher.
This shows that TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occasionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
For example, the distribution of comparison between PB1 and PB2 trees has a mean $\dtmrca$$\approx0.4$, indicating that the sum of $\dtmrca$ values from PB1-PB1' and PB2-PB2' comparisons (prime (') indicates an independent MCMC chain) are 40\% of twice the $\dtmrca$ value from the PB1-PB2 comparison, but this value can range anywhere between $\approx$0.15 and $\approx$0.75.
Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort among themselves unless both reassorting segments have similar TMRCAs, i.e.\ we observe isolation by temporal distance.
In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.

@@

Linkage disequilibrium (LD) is a measure of non-random association between pairs of alleles at different polymorphic loci within a population.
To estimate LD no other information, except for haplotype and allele frequencies at polymorphic amino acid sites, is required.
Although not an external validation of previous phylogenetic results, we observe significantly greater LD values between PB1, PB2 and HA than between other pairs of segments (Figure \ref{segmentLD}).
This suggests that PB1, PB2 and HA segments possess a considerable number of co-assorting alleles, which upon closer inspection are associated with either Vic or Yam lineage segments.
We conclude that Victoria and Yamagata lineages of PB1, PB2 and HA have accumulated lineage-specific amino acid substitutions.
Of the amino acid sites that exhibit high LD on PB1, PB2 and HA proteins, there are 5 sites on PB1, 8 on PB2 and 13 on HA proteins which form a network of sites exhibiting high LD (Figures \ref{ChiGenome} and \ref{ChiCore}).
Figure \ref{ChiGenome} also shows that the number of variable amino acid sites in PB1 and PB2 proteins is comparable to most other proteins investigated (see Methods for cut-offs used to exclude sites from LD analysis).
These sites define the split between Vic and Yam lineages within PB1, PB2 and HA segments.
In addition, there are sites on PB1, PB2 and HA proteins which also show high, albeit smaller, LD which correspond to sites which have undergone amino acid replacements some time after the Vic/Yam split.

@@

When comparing the mean tip-tip TMRCA deviations we find evidence of isolation by temporal distance - when reassortments occur between PB1, PB2 and HA segments those events tend to involve branches that have similar TMRCAs.
Other segment pairs, with the exception of MP-NA, seem to show little signal of this and are able to reassort with branches that have been evolving under temporally distant genomic backgrounds.
In addition, most strains with mixed-lineage PB1-PB2-HA complexes occurred in the early years of the Vic--Yam split, when the two lineages were presumably much more similar at the nucleotide and amino acid levels.
The most recent influenza B viruses with B/Waikato/6/2005-like genome constellations possessed mixed-lineage PB1-PB2-HA complexes and circulated for only 5 months before presumably going extinct.

@@

\subsection*{Causes of co-dependence between PB1-PB2-HA}
Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, \textit{e.g.} PB1-PB2-PA and HA-NA \citep{mccullers2004}.
Though it would be easy to explain co-adaptation between these segments by referring to their functional roles \textit{e.g.} PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities, our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.

It is clear that PB1-PB2-HA segments of Vic and Yam lineages do not preferentially reassort together: there have been at least 4 sampled mixed-lineage PB1-PB2-HA complex constellations which did not become fixed in the population and our estimates of the number of reassortments do not differ significantly between all segments (Figure \ref{SPRdistances}).
The latter is in line with recent experiments in influenza A that have shown that reassortment between segments differing by a single synonymous difference is highly efficient \citep{marshall2013}.
In the absence of clear functional explanations for why PB1, PB2 and HA should be co-adapted we offer several alternatives.

@@

Explaining the co-dependence of PB1+2 and HA segments is more difficult.
Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \citep{bergeron2010,fulvini2011}.
A previously used technique for producing vaccine seed strains involved selecting for HA-NA reassortants, which often resulted in PB1-HA-NA reassortants, the second most frequent class of reassortants \citep{bergeron2010,fulvini2011}.
Recent experiments have suggested that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \citep{cobbin2013}.
Additional evidence for a relationship between PB1 and HA segments in influenza A viruses is given by previous influenza pandemics, which were caused by avian-human influenza A virus reassortants.
It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments derived from avian influenza A viruses \citep{kawaoka1989}.
However, there have also been reassortant influenza A viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \citep{gregory2002} and H1N1/09 in 2009 \citep{smith2009}.

@@

Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \citep{chen2006,obenauer2006}.
This is easiest to interpret as frequency-dependent balancing selection acting to preserve antigenic diversity \citep{worobey2014}.
If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.

It is possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
In this case PB1, PB2 and HA segments of Vic and Yam lineage accumulate substitutions that improve their ability to co-operate with segments of the same lineage and worse at interacting segments of a different lineage.
This process, termed mutation-driven co-evolution \citep{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \citep{lee2008}.
It is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering a potential explanation for why the budding Victoria lineage segments were not homogenized via reassortment with Yamagata lineage segment in the early years of the split between the two lineages. 
The restriction to eastern Asia would presumably also give sufficient time for PB1, PB2 and HA segments to co-evolve together.
Similarly to the previous scenario, however, mutation-driven co-evolution would require an association of some kind between PB1+2 and HA segments to explain the low observed frequency and poor fitness of PB1+2/HA reassortants.
In addition, we see no reason why these three particular segments, and not the rest of the genome, would become co-adapted to each other in geographic isolation.

\subsection*{Suggested experiments}
The association between Victoria and Yamagata lineage PB1-PB2-HA complexes should be relatively straightforward to test in the lab.
Using previously developed plasmid systems \citep{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
We predict that artificially produced viruses with mixed-lineage PB1-PB2-HA complexes will have reduced fitness when compared to viruses with pure-lineage, i.e.\ entirely Vic or entirely Yam, PB1-PB2-HA complexes.
In addition, we expect the relationship between Vic and Yam lineage PB1, PB2 and HA segments to be dependent on date of segment isolation, as viruses with mixed-lineage PB1-PB2-HA complexes isolated earlier should perform better than viruses with PB1-PB2-HA segments isolated more recently.

@@

However, we also see that epistatic effects might interfere with fitness measurements, if for example non-PB1-PB2-HA segments are also temporally mismatched.
Ideally, the co-adaptation would be easier to understand by referring to the structures of PB1 and PB2 proteins, as the link between these would be intuitive.
We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins (Figure \ref{ChiCore}) that fall within the regions that form contacts within the influenza B polymerase heterotrimer \citep{sugiyama2009}, suggesting more subtle roles for sites we have identified.

\subsection*{The future of influenza B viruses}
We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \citep{turner2005}.
In this context, we see three potential paths of evolution for influenza B viruses.
More segments could be recruited into the two currently circulating co-adapted segment complexes (PB1, PB2 and HA segments being the genomic speciation islands), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
This is the speciation scenario.
The two alternatives to this are the `\textit{status quo}' model, where the influenza B genome continues to be homogenized via gene flow with the exception of PB1, PB2 and HA segments and the extinction scenario, whereby one of the two PB1-PB2-HA complexes goes extinct, marking the return of single-strain dynamics in the influenza B virus population.
